---
title: "Introduction to jamsession"
author: "James M. Ward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to jamsession}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", warning=FALSE);
#library(jamsession);
```

# Introduction to jamsession

The jamsession package and functions are motived primarily to make your
(my) analysis workflows as efficient as possible. They start with the simple
idea of storing certain information in a central location, partly because it
makes it easier to find months or years later. This workflow also works well
if more often than not, you have multiple R sessions open and active, since
these functions add the project name to the R prompt.

The R session and object files are versioned with the date, which has
been an appropriate level of granularity, as well as being useful information
when returning to work on past projects.

That said, this package
mainly focuses on three aspects of R analysis work:

### R sessions

This package stores and loads R sessions via .RData files from one
stored location. These methods are useful are when analyzing
data on different projects, where one may be asked to review
data or results from projects years after the initial analysis.
The methods here also store the R command history, which can be helpful
when picking up where one left off.

Saving or loading an R session also sets the R prompt to the project name,
which is particularly helpful when multiple R sessions are open an active
at once.

### R objects

This package also stored individual R objects in their own central location,
in order to make it efficient to save and load objects in any R session,
regardless the project or current working directory.

### R functions

This package allows a clean mechanism of using a helper R script with functions
to load and reload for active analysis use. Doing so minimizes copy-paste into
active consoles, including via Rstudio, but also keeps function code in their
own environment instead of the global environment. When stored in the global
environment, the function code is saved along with the R session, which may not
be desirable especially when using a large set of helper functions.

## Motivations

Some motivations for this package:

- make it efficient to find past R sessions and R objects by name, without
the need to traverse down a nested file directory structure
- make stored sessions and objects versioned by date, rather than using
only the project or object name, and not needing to increment a version number
every time a file is saved
- provide a quick way to save and load R objects across different R sessions,
to share useful objects in a structured, repeatable way
- provide a clean method to load R functions from a separate file, to avoid
having to copy-paste functions inside an R script, and to avoid R functions
becoming part of the environment


## Basic R session workflow

When starting a new project in a new R console, it is helpful to initiate
the environment with the project name, by saving a new session. By default,
`session` will be defined during this step, and the R prompt will
change to display the `session`.

> Note that we add the argument `sessions_path=tempdir()` only for this
vignette, in typical use this argument is not used, because R
sessions are stored in one location.

```{r, jamsession_1}
library(jamsession);
some_object <- data.frame(A=LETTERS[1:10]);
save_jamsession("jamsessionVignette",
   sessions_path=tempdir());
print(session);
```

You may then do some work and wish to save your session:

```{r, jamsession_2}
x <- rnorm(90);
y <- matrix(ncol=9, data=x);
## Usually only run like this:
# save_jamsession();

## For this vignette we add sessions_path
save_jamsession(sessions_path=tempdir());
```

You can later start R and pick up where you left off, which is simulated
here by removing these objects, then loading the session:

```{r, jamsession_3}
rm(y);
load_jamsession("jamsessionVignette",
   sessions_path=tempdir());
print(head(y));
```

## Saving R objects

If you want to save an object, it is best to name it something specific before
saving it. The call the `saveObject()` function with the name of the object:

> Note that we add the argument `objects_path=tempdir()` only for this
vignette, in typical use this argument is not used, because R
objects are stored in one location.

```{r, objects_1}
jamsessionVignetteY <- t;
save_object("jamsessionVignetteY",
   objects_path=tempdir());
```

You can then load this object from any R console using `load_object()`, simulated
here by removing the object, then loading it from the saved object file:

```{r, objects_2}
rm(jamsessionVignetteY);
load_object("jamsessionVignetteY",
   objects_path=tempdir());
print(head(jamsessionVignetteY));
```


## Loading R functions

R is a functional programming language, which makes it especially useful
and powerful to write and re-use custom functions. Most R analysts have built
up a long list of useful helper functions, either for re-use across all
their projects, or written specifically per project. Either way, these
functions need to be loaded into the R workspace, and very likely will need
to be refreshed as the functions are updated over time.

We demonstrate having a utility script by creating a temporary script, then
using `refresh_functions()`.

```{r, printfun_1}
## Create a simple R script function fun()
r_script <- "
#' Simple print function
#' 
#' This function simply runs print()
#'
#' @param x any R object that can be printed.
#' @param ... additional arguments are passed to print().
#'
#' @export
printfun <- function(x, ...) {print(x, ...)};
";

## Save to a file in tempdir
fn_file <- file.path(tempdir(),
   "jamvignette_functions.R");
## Save the R script to a file
cat(r_script, file=fn_file);
```

Now the file has been saved under the name "jamsession". We can load this
file with `refresh_functions()`.

> For this vignette, we include argument `functions_path=tempdir()`,
but this argument is usually not necessary, when functions are
saved to a common path. Run `jamvignette_paths()` to see or set
the expected paths.

```{r, printfun_2}
## Load the R script
refresh_functions("jamvignette",
   functions_path=tempdir());

## search() shows the environments including jamvignette
print(search());

## find() to show where printfun is located
find("printfun")
```

Lastly, we can run the `printfun()` function to show it works.

```{r, printfun_3}
printfun(LETTERS[1:10]);
```
